State of the Practice for Medical Imaging Software

### 2021-11-24 StateOfPractice/Papers/MedImageSoft_SOP/MedImageSoft_SOP_AcademicRadiology.tex
Edits to MI SOP up to (but not including) Section 5 (Interviews with Developers)

%\documentclass[preprint,12pt,authoryear]{elsarticle}
\documentclass[3p, 12pt,authoryear]{elsarticle}

@@

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  Below we list the 10 qualities of interest for this study,
as summarized in \citet{SmithEtAl2021-QDef}.  The order of the qualities follows
the order used in \cite{GhezziEtAl2003}, which puts related qualities (like
correctness and reliability) together.  Moreover, the order is roughly the same
as the order qualities are considered in practice.  [*** To be completed - not
currently in sync with QDef paper, QDef paper not completely done ***]

@@

In this project, the first nine software qualities mentioned in Section
\ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
packages ($n = 29$) are compared for each of the $m$ criteria. The software are
evaluated with the grading template in Section~\ref{sec_grading_template}, which
includes a subjective score from $1$ to $10$ for each quality for each package.
For each quality, for a pair of packages $i$ and $j$, with $\mathit{score}_i >=
\mathit{score}_j$, the difference between the two scores is $\mathit{diff_{ij}}
= \mathit{score}_i - \mathit{score}_j$. The mapping between $\mathit{diff_{ij}}$
(which can vary between 0 and 9) and the values in $A_{ij}$ (which can vary
between 1 and 9) is as follows:

@@

We grade the selected software using a template (Section
\ref{sec_grading_template}) and a specific empirical method (Section
\ref{sec_empirical_measurements}). Some technical details for the measurements
are in Section \ref{sec_technical_details}.  We do not measure
\textit{reproducibility} with the grading template, but discuss it with the
developers by interviews.

The full grading template can be found in Appendix \ref{ap_grading_template}.
The template contains 103 questions that we use for grading software products.
Figure \ref{fg_grading_template_example} shows an example of this grading
template.

@@

We use the first section of the template to collect general information, such as
the name, purpose, platform, programming language, publications about the
software, the first release and the most recent change date, website, source
code repository of the product, etc. Information in this section helps us
understand the projects better and may be helpful for further analysis, but it
does not directly affect the grading scores.

We designed the following nine sections in the template for the nine software
qualities mentioned in Section~\ref{sec_software_quality}. For each quality, we
ask several questions and the typical answers are among the collection of
``yes'', ``no'', ``n/a'', ``unclear'', a number, a string, a date, a set of
strings, etc. Each quality needs an overall score between 1 and 10 based on all
the previous questions. For some qualities, we perform surface measurements,
which allow us to measure all packages with reasonable efforts. The surface
measurements reveal some traits of a underlying quality, but may not fully
represent it.

@@

\item \textbf{Installability} We check the existence and quality of installation
instructions. The user experience is also an important factor, such as the ease
to follow the instructions, number of steps, automation tools, and the
prerequisite steps for the installation. If any problem interrupts the process
of installation or uninstallation, we give a lower score to this quality. We
also record the Operating System (OS) for the installation test and whether we
can verify the installation.

\item \textbf{Correctness \& Verifiability} For \textit{correctness}, we check
the projects to identify the techniques to ensure this quality, such as literate
programming, automated testing, symbolic execution, model checking, unit tests,
etc. We also examine whether the projects use continuous integration and
continuous delivery (CI/CD). For \textit{verifiability}, we go through the
documents of the projects to check the requirements specifications, theory
manuals, and getting started tutorials. If a getting started tutorial exists and
provides expected results, we follow it and check if the outputs match.

\item \textbf{Surface Reliability} We check whether the software breaks during
the installation and the operations following a getting started tutorials,
whether there are descriptive error messages, and if we can recover the process
after an error. We use the software to load damaged images during the assessment
to \textit{surface robustness}, and lower its score for \textit{surface
reliability} if a software product ``break" in this process.

\item \textbf{Surface Robustness} We check how the software handles
unexpected/unanticipated input. For example, we prepare broken image files for
MI software packages with the function to load image files. We use a text file
(.txt) with a modified extension name (.dcm) as an unexpected/unanticipated
input. We load a few correct input files to ensure the function is working
correctly before testing with the unexpected/unanticipated ones.

\item \textbf{Surface Usability} We examine the documentation of the projects,
and we consider software with a getting started tutorial and a user manual
easier to use. Meanwhile, we check if users have any channels to get support. We
also record our impressions and user experiences when testing the software.
Easy-to-use graphical user interfaces give us a better experience, which leads
to better scores.

\item \textbf{Maintainability} We search the projects' documents and identify
the process of contributing and reviewing code. We believe that the artifacts of
a project - including source code, documents, building scripts, etc. - can
significantly affect its  \textit{maintainability}. Thus we check each project
for its artifacts, such as API documentation, bug tracker, release notes, test
cases, build files, version control, etc. We also check the tools supporting
issue tracking and version control, the percentages of closed issues, and the
proportion of comment lines in code.

\item \textbf{Reusability} We count the total number of code files for each
project. Projects with a large number of components provide more choices to
reuse. Furthermore, well-modularized code, which tends to have smaller parts in
separated files, is typically easier to reuse. Thus, we consider the projects
with more code files and less code lines per file to be more reusable. We use a
command-line tool \textit{scc} to count the number of  text-based files and LOC
for all projects. We also decide that the projects with API documentation can
deliver better \textit{Reusability}.

\item \textbf{Surface Understandability} We randomly examine 10 code files. We
check the code’s style within each file, such as whether the identifiers,
parameters, indentation, and formatting are consistent, whether the constants
(other than 0 and 1) are hardcoded, and whether the developers modularized the
code. We also check the descriptive information for the code, such as documents
mentioning the coding standard, the comments in the code, and the descriptions
or links for algorithms in the code.

\item \textbf{Visibility/Transparency} To measure this quality, we check the
existing documents to find out whether the software development process and
current status of a project are visible and transparent. We examine the
development process, current status, development environment, and release notes
for each project. If any information is missing or poorly conveyed, the
\textit{visibility/transparency} will be lower.
\end{itemize}

For some qualities, the empirical measurements also affect the score. We use
tools to extract information from the source code repositories. For projects
held on GitHub, we manually collect additional metrics, such as the stars of the
GitHub repository, and the numbers of open and closed pull requests. Section
\ref{sec_empirical_measurements} presents more details about the empirical
measurements.

We use two command-line tools for the empirical measurements. One is
\textit{GitStats} that generates statistics for git repositories and displays
outputs in the form of web pages \citep{Gieniusz2019}; the other one is Sloc Cloc
and Code (as known as \textit{scc}) \citep{Boyter2021}, aiming to count the lines
of code, comments, etc.

@@

Additionally, we manually collect information for projects held on GitHub, such
as the numbers of stars, forks, people watching this repository, open pull
requests, closed pull requests, and the number of months a repository has been
on GitHub. A git repository can have a creation date much earlier than the first
day on GitHub. For example, the developers created the git repository of
\textit{3D Slicer} in 2002, but did not upload a copy of it to GitHub until
2020. We get the creation date of the GitHub copy by using API
\textit{https://api.github.com/repos/{:owner}/{:repository}} (e.g.,
\hyperlink{https://api.github.com/repos/slicer/slicer}{https://api.github.com/repos/slicer/slicer}).
In the response, the value of ``created\_at" is what we want. The number of
months a repository has been on GitHub helps us understand the average change of
metrics over time, e.g., the average new stars per month.

These empirical measurements help us from two aspects. Firstly, they help us
with getting a project overview faster and more accurately. For example, the
number of commits over the last 12 months shows how active this project has
been, and the number of stars and forks may reveal its popularity. Secondly, the
results may affect our decisions regarding the grading scores for some software
qualities. For example, if the percentage of comment lines is low, we
double-check the \textit{understandability} of the code; if the ratio of open
versus closed pull requests is high, we pay more attention to the
\textit{maintainability}.

As in \citet{SmithEtAl2016}, Virtual machines (VMs) are used to provide an
optimal testing environments for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors. When
filling in the measurement template spreadsheet, the the details for each VM
should be noted, including hypervisor and operating system version.

We spend about two hours grading each package, unless we find technical issues
and need more time to resolve them. In most of the situation, we finish all the
measurements for one software on the same day.

Then we followed the steps in Section \ref{sec_grading_software} to measure and
grade the software. 27 out of the 29 packages are compatible with two or three
different OS such as Windows, macOS, and Linux, and 5 of them are browser-based,
making them platform-independent. However, in the interest of time, we only
performed the measurements for each project by installing it on one of the
platforms, most likely Windows.

@@

This section introduces our interview questions (Section
\ref{sec_interview_questions}), method of selecting interviewees (Section
\ref{sec_interviewee_selection}), and interview process (Section
\ref{sec_interview_process}).

We designed a list of 20 questions to guide our interviews, which can be found
in Section \ref{ch_interview} and Appendix \ref{ap_interview}.

Some questions are about the background of the software, the development teams,
the interviewees, and how they organize the projects. We also ask about their
understandings of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found or will try. We also discuss
the importance and current situations of documentation. A few questions are
about specific software qualities, such as \textit{maintainability},
\textit{understandability}, \textit{usability}, and \textit{reproducibility}.

The interviews are semi-structured based on the question list; we ask follow-up
questions when necessary. Based on our experience, the interviewees usually
bring up some exciting ideas that we did not expect, and it is worth expanding
on these topics.

For a software list with a length of roughly 30, we aim to interview about ten
development teams. Interviewing multiple individuals from each team gives us
more comprehensive information, but a single engineer well-knowing the project
is also sufficient.

Ideally, we select projects after the grading measurements and prefer the ones
with higher overall scores. However, if we do not find enough participants, we
reach all teams on the list. As mentioned in Section
\ref{sec_apply_to_mi_interviews}, when we applied this process to the MI domain,
we eventually contacted all teams.

We try to find the contacts of the teams on the projects' websites, such as the
official web pages, repositories, publications, and bio pages of the teams'
institutions. Then, we send at most two emails to one contact asking for its
participation before receiving any replies. We operate the invitation according
to our ethics approval, such as the one in Appendix \ref{ap_ethics}. For
example, we ask for participants' consent before interviewing them, recording
the conversation, or including it in our report.

Before contacting any interviewee candidate, we need to receive ethics clearance
from the McMaster University Research Ethics Board. Since the members of the
development teams are usually around the world, we organize these interviews as
virtual meetings online with \hyperlink{https://zoom.us/}{Zoom}. After receiving
consent from the interviewees, we also record and transcribe our discussions.

We received ethics clearance from the McMaster University Research Ethics Board
(Appendix \ref{ap_ethics}). Going through the interview process in Section
\ref{sec_interview_methods}, we contacted all of the 29 teams. Members from
eight teams responded and agreed to participate. As a result, we interviewed
nine developers and architects from the eight teams.

@@

As discussed in Section~\ref{sec_grading_software}, we use a grading template
and a empirical method to measure the selected software. We applied this step to
the MI domain (Section~\ref{sec_applying_method_grading}). This section shows
the summary of the measurement results. The detailed data can be found in the
repository
\hyperlink{https://data.mendeley.com/datasets/k3pcdvdzj2/1}{https://data.mendeley.com/datasets/k3pcdvdzj2/1}.
This section contains part of the answers to \hyperlink{rq5}{RQ5}.

Table \ref{tab_final_list} shows the 29 software packages that we measured,
along with some summary data collected in the year 2020. As mentioned in Section
\ref{sec_grading_template}, we used \textit{scc} (Section
\ref{sec_empirical_measurements}) to count the Lines of Code (LOC), excluding
the comment and blank lines. We arrange the items in the descending order of the
LOC. We found the initial release dates (Rlsd) in their documents for most
projects and marked the two unknown dates with ``?". We used the date of the
latest change to each code repository to decide the latest update. We found out
funding information (Fnd) for only eight projects.

We counted the number of contributors (NOC). We considered anyone who made at
least one accepted commit to the source code as a contributor. Thus, the NOC is
not usually the same as the number of long-term members. Many of these projects
received change requests and code from the community, such as pull requests and
git commits on GitHub.

Table \ref{tab_final_list} also shows the supported OS for each software
package. Twenty-five of them could work on all three OSs: Windows (W), macOS
(M), and Linux (L). However, there was a significant difference in the
philosophy to achieve cross-platform compatibility. Most of them were native
software products, but five were naturally platform-independent web
applications, as shown in column ``Web".

\begin{table}[ht]
\centering
\begin{tabular}{llllllllll}
\hline
\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
 &  &  &  &  &  & W & M & L &  \\ \hline
ParaView \citep{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & X &  &  \\
OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & X &  &  \\
3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
3DimViewer \citep{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & X & X & X & X \\
DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & X & X & X & X \\
dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & X & X &  &  \\
Fiji \citep{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
ImageJ \citep{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \hline
\end{tabular}
\caption{\label{tab_final_list}Final software list}
\end{table}

Most of the projects used more than one programming language, including a
primary language that the developers used the most. Figure \ref{fig_language}
shows the primary languages versus the number of projects using them.

% temporarily comment out

% \begin{figure}[ht]
% \centering
% \begin{tikzpicture}
% \centering
% \begin{axis}
% [
% ybar,
% height=8cm,
% width=12cm,
% ylabel={Number of projects},
% xlabel={\ Primary language},
% symbolic x coords={C++, JavaScript, Java, C, Python, Pascal, Matlab},
% xtick=data,
% nodes near coords,
% nodes near coords align={vertical},
% ]
% \addplot[black,fill=blue!55!black] coordinates {(C++,11) (JavaScript,6) (Java,4) (C,3) (Python,3) (Pascal,1) (Matlab,1) };
% \end{axis}  
% \end{tikzpicture}
% \caption{\label{fig_language}Primary languages versus number of projects using them}
% \end{figure}

We failed to install \textit{DICOM Viewer}, so we could not test its
\textit{surface reliability} and \textit{surface robustness}. We kept this
software on our list because the other seven qualities do not rely on a
successful installation. Besides, the \textit{DICOM Viewer} team built it as a
plugin software for NextCloud
(\hyperlink{https://apps.nextcloud.com/}{https://apps.nextcloud.com/}) platform,
which was a unique choice we had not seen before. We wanted to keep it to enrich
the diversity.

@@

Figure \ref{fg_installability_scores} lists the scores of
\textit{installability}.

@@

We found installation instructions for 16 projects. Among the ones without
instructions, \textit{BioImage Suite Web} and \textit{Slice:Drop} are web
applications with online versions to use, thus they do not need installation.
Installing 10 of the projects required extra dependencies. Five of them are the
web applications in Table \ref{tab_final_list}, and depended on a browser;
\textit{dwv}, \textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies
to build; \textit{ImageJ} and	\textit{Fiji} needed an unzip tool;
\textit{MatrixUser} was based on Matlab; \textit{DICOM Viewer} needed to work on
a Nextcloud platform.

>>>>

%\documentclass[preprint,12pt,authoryear]{elsarticle}
%\documentclass[3p, 12pt,authoryear]{elsarticle}
\documentclass[final, 3p, times, authoryear]{elsarticle}

@@

Quality is defined as a measure of the excellence or worth of an entity.  As is
common practice, we do not think of quality as a single measure, but rather as a
set of measures.  That is, quality is a collection of different qualities, often
called ``ilities.''  Below we list the 10 qualities of interest for this study,
as summarized in \citet{SmithEtAl2021-QDef}.  The order of the qualities follows
the order used in \cite{GhezziEtAl2003}, which puts related qualities (like
correctness and reliability) together.  Moreover, the order is roughly the same
as the order qualities are considered in practice.  \wss{To be completed - not
currently in sync with QDef paper, QDef paper not completely done.}

@@

In this project, the first nine software qualities mentioned in Section
\ref{sec_software_quality} are the criteria ($m = 9$), while 29 software
packages ($n = 29$) are compared for each of the $m$ criteria. The software are
evaluated with the grading template in Section~\ref{sec_grading_software}, which
includes a subjective score from $1$ to $10$ for each quality for each package.
For each quality, for a pair of packages $i$ and $j$, with $\mathit{score}_i >=
\mathit{score}_j$, the difference between the two scores is $\mathit{diff_{ij}}
= \mathit{score}_i - \mathit{score}_j$. The mapping between $\mathit{diff_{ij}}$
(which can vary between 0 and 9) and the values in $A_{ij}$ (which can vary
between 1 and 9) is as follows:

@@

We grade the selected software using the measurement template summarized
in~\citet{SmithEtAl2021}.  The template summarizes measures of the qualities
listed in Section~\ref{sec_software_quality}, except for reproducibility, which
is assessed through the developer interviews
(Section~\ref{sec_interview_methods}). For each software package, we fill-in the
template questions. This process can take between 1 to 4 hours for each package.
Project developers can be contacted for help regarding installation, if
necessary, but a cap of about 2 hours is imposed on the
installation process, to keep the overall measurement time feasible. An excerpt
of the spreadsheet is shown in Figure~\ref{fg_grading_template_example}.
\wss{The Figure should be updated to be easier to read.}  A column is included
to for each measured software package.

@@

The full template consists of 108 questions categorized under 9 qualities.  The
questions were designed to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. The measures are grouped under headings for
each quality, and one for summary information. The summary information (shown in
Figure~\ref{fg_grading_template_example}) is the first section of the template.
This section summarizes general information, such as the software name, purpose,
platform, programming language, publications about the software, the first
release and the most recent change date, website, source code repository of the
product, number of developers, etc.  We follow the definitions given by
\citet{GewaltigAndCannon2012} for the software categories.  Public means
software intended for public use.  Private means software aimed only at a
specific group, while the concept category is used for software written simply
to demonstrate algorithms or concepts. The three categories of development
models are (open source, free-ware and commercial) are discussed in
Section~\ref{sec_software_categories}.  Information in the summary section sets
the context for the project, but it does not directly affect the grading scores.

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. Each quality is assigned an overall
score, between 1 and 10, based on all the previous questions.  Several of the
qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that usability was considered by the
developers.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and documentation of expected user
characteristics.  Below is a summary of how each quality is measured.

@@

\item \textbf{Installability} We assess the following: 
\begin{inparaenum}[i)]
    \item existence and quality of installation instructions;
    \item the quality of the user experience via the ease of following
    instructions, number of steps, automation tools; and,
    \item whether there is a means to verify the installation.
\end{inparaenum}
If any problem interrupts the process of installation or uninstallation, we give
a lower score. We also record the Operating System (OS) used for the
installation test.

\item \textbf{Correctness \& Verifiability} We check each project to identify
any techniques used to ensure this quality, such as literate programming,
automated testing, symbolic execution, model checking, unit tests, etc. We also
examine whether the projects use continuous integration and continuous delivery
(CI/CD). For verifiability, we go through the documents of the projects to check
for the presence of requirements specifications, theory manuals, and getting
started tutorials. If a getting started tutorial exists and provides expected
results, we follow it to check the correctness of the output.

\item \textbf{Surface Reliability} We check the following: 
\begin{inparaenum}
    \item whether the software breaks during installation;
    \item the operation of the software following the getting started tutorial (if present);
    \item whether the error messages are descriptive; and,
    \item whether we can recover the process after an error.
\end{inparaenum}

\item \textbf{Surface Robustness} We check how the software handles
unexpected/unanticipated input. For example, we prepare broken image files for
MI software packages that load image files. We use a text file (.txt) with a
modified extension name (.dcm) as an unexpected/unanticipated input. We load a
few correct input files to ensure the function is working correctly before
testing the unexpected/unanticipated ones.

\item \textbf{Surface Usability} We examine the project's documentation,
checking for the presence of getting started tutorials and/or a user manual. We
also check whether users have channels to request support, such as an e-mail
address, or issue tracker. Our impressions of usability are based on our
interaction with the software during testing.  In general, an easy-to-use
graphical user interface scores higher.

\item \textbf{Maintainability} We believe that the artifacts of a project,
including source code, documents, and building scripts, significantly influence
its maintainability. Thus we check each project for the presence of such
artifacts as API documentation, bug tracker information, release notes, test
cases, and build files. We also check for the use of tools supporting issue
tracking and version control, the percentages of closed issues, and the
proportion of comment lines in the code.

\item \textbf{Reusability} We count the total number of code files for each
project. Projects with a large number of components potentially provide more
choices for reuse. Furthermore, well-modularized code, which tends to have
smaller parts in separate files, is typically easier to reuse. Thus, we consider
the projects with more code files and less Lines of Code (LOC) per file to be
more reusable. We also consider projects with API documentation as delivering
better reusability.

\item \textbf{Surface Understandability} Given that time is a constraint, we
cannot look at all code files for each project; therefore, we randomly examine
10 code files for their understandability. We check the code’s style within each
file, such as whether the identifiers, parameters, indentation, and formatting
are consistent, whether the constants (other than 0 and 1) are hardcoded, and
whether the code is modularized. We also check the descriptive information for
the code, such as documents mentioning the coding standard, the comments in the
code, and the descriptions or links for details on algorithms in the code.

\item \textbf{Visibility/Transparency} To measure this quality, we check the
existing documents to find whether the software development process and
current status of a project are visible and transparent. We examine the
development process, current status, development environment, and release notes
for each project.
\end{itemize}

As part of filling in the measurement template, we use freeware tools to collect
repository related data. \href{https://github.com/tomgi/git_stats}{GitStats}
\citep{Gieniusz2019} is used to measure the number of binary files as well as
the number of added and deleted lines in a repository. This tool is also used to
measure the number of commits over different intervals of time.
\href{https://github.com/boyter/scc}{Sloc Cloc and Code (scc)}
\citep{Boyter2021} is used to measure the number of text based files as well as
the number of total, code, comment, and blank lines in a repository.

@@

For projects on GitHub we manually collect additional information, such as the
numbers of stars, forks, people watching this repository, and open pull requests
on GitHub. A git repository can have a creation date much earlier than the first
pull, closed pull requests, and the number of months a repository has been on
GitHub. For example, the developers created the git repository for \textit{3D
Slicer} in 2002, but did not upload a copy of it to GitHub until 2020. Some
GitHub data can be found using its the GitHub Application Program Interface
(API) via the following url:
\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
[repository] are replaced by the repo specific values. The number of months a
repository has been on GitHub helps us understand the average change of metrics
over time, like the average new stars per month.

The repository measures help us in wo ways. Firstly, they help us with get a
fast and accurate project overview. For example, the number of commits over the
last 12 months shows how active a project has been, and the number of stars and
forks may reveal its popularity. Secondly, the results may affect our decisions
regarding the grading scores for some software qualities. For example, if the
percentage of comment lines is low, we double-check the
\textit{understandability} of the code; if the ratio of open versus closed pull
requests is high, we pay more attention to \textit{maintainability}.

As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
optimal testing environments for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors. When
filling in the measurement template spreadsheet, the the details for each VM
should be noted, including hypervisor and operating system version.

When grading the software, we found 27 out of the 29 packages are compatible
with two or three different OS such as Windows, macOS, and Linux, and 5 of them
are browser-based, making them platform-independent. However, in the interest of
time, we only performed the measurements for each project by installing it on
one of the platforms.  When it was an option, we selected Windows as the host
OS.

@@

We designed a list of 20 questions to guide our interviews, which can be found
in \citet{SmithEtAl2021}. Some questions are about the background of
the software, the development teams, the interviewees, and how they organize the
projects. We also ask about the developer's understandings of the users. Some
questions focus on the current and past difficulties, and the solutions the team
has found, or will try. We also discuss the importance and current situations of
documentation. A few questions are about specific software qualities, such as
maintainability, understandability, usability, and reproducibility. The
interviews are semi-structured based on the question list; we ask follow-up
questions when necessary. Based on our experience, the interviewees usually
bring up some exciting ideas that we did not expect, and it is worth expanding
on these topics.

We sent out interview requests to all 29 projects, with 9 projects responding.
In some cases multiple developers from the same project agreed to be
interviewed. We found contact information from projects websites, code
repository, publications, and from biographic pages at the teams’ institutions.
Meetings were held on-line using either Zoom or Teams, which facilitates
recording and automatic transcription of the meetings.  The interview process
presented here was approved by the McMaster University Research Ethics Board
under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

Table \ref{tab_final_list} shows the 29 software packages that we measured,
along with summary data collected in the year 2020. We arrange the items in
descending order of LOC. We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?". We used the date of the
latest change to each code repository to decide the latest update. We found
funding information (Fnd) for only eight projects.  For the number of
contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  For the supported OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available in \citet{Dong2021-Data}.

\begin{table}[ht]
\centering
\begin{tabular}{p{6cm}lllllllll}
\hline
\multirow{2}{*}{Software} & \multirow{2}{*}{Rlsd} & \multirow{2}{*}{Updated} & \multirow{2}{*}{Fnd} & \multirow{2}{*}{NOC} & \multirow{2}{*}{LOC} & \multicolumn{3}{c}{OS} & \multirow{2}{*}{Web} \\ \cline{7-9}
 &  &  &  &  &  & W & M & L &  \\ \hline
ParaView \citep{Ahrens2005} & 2002 & 2020-10 & X & 100 & 886326 & X & X & X & X \\
Gwyddion \citep{Nevcas2012} & 2004 & 2020-11 &  & 38 & 643427 & X & X & X &  \\
Horos \citep{horosproject2020} & ? & 2020-04 &  & 21 & 561617 &  & X &  &  \\
OsiriX Lite \citep{PixmeoSARL2019} & 2004 & 2019-11 &  & 9 & 544304 &  & X &  &  \\
3D Slicer \citep{Kikinis2014} & 1998 & 2020-08 & X & 100 & 501451 & X & X & X &  \\
Drishti \citep{Limaye2012} & 2012 & 2020-08 &  & 1 & 268168 & X & X & X &  \\
Ginkgo CADx \citep{Wollny2020} & 2010 & 2019-05 &  & 3 & 257144 & X & X & X &  \\
GATE \citep{Jan2004} & 2011 & 2020-10 &  & 45 & 207122 &  & X & X &  \\
3DimViewer \citep{TESCAN2020} & ? & 2020-03 & X & 3 & 178065 & X & X &  &  \\
medInria \citep{Fillard2012} & 2009 & 2020-11 &  & 21 & 148924 & X & X & X &  \\
BioImage Suite Web \citep{Papademetris2005} & 2018 & 2020-10 & X & 13 & 139699 & X & X & X & X \\
Weasis \citep{Roduit2021} & 2010 & 2020-08 &  & 8 & 123272 & X & X & X &  \\
AMIDE \citep{Loening2017} & 2006 & 2017-01 &  & 4 & 102827 & X & X & X &  \\
XMedCon \citep{Nolf2003} & 2000 & 2020-08 &  & 2 & 96767 & X & X & X &  \\
ITK-SNAP \citep{Yushkevich2006} & 2006 & 2020-06 & X & 13 & 88530 & X & X & X &  \\
Papaya \citep{UTHSCSA2019} & 2012 & 2019-05 &  & 9 & 71831 & X & X & X &  \\
OHIF Viewer \citep{Ziegler2020} & 2015 & 2020-10 &  & 76 & 63951 & X & X & X & X \\
SMILI \citep{Chandra2018} & 2014 & 2020-06 &  & 9 & 62626 & X & X & X &  \\
INVESALIUS 3 \citep{Amorim2015} & 2009 & 2020-09 &  & 10 & 48605 & X & X & X &  \\
dwv \citep{Martelli2021} & 2012 & 2020-09 &  & 22 & 47815 & X & X & X & X \\
DICOM Viewer \citep{Afsar2021} & 2018 & 2020-04 & X & 5 & 30761 & X & X & X &  \\
MicroView \citep{ParallaxInnovations2020} & 2015 & 2020-08 &  & 2 & 27470 & X & X & X &  \\
MatrixUser \citep{Liu2016} & 2013 & 2018-07 &  & 1 & 23121 & X & X & X &  \\
Slice:Drop \citep{Haehn2013} & 2012 & 2020-04 &  & 3 & 19020 & X & X & X & X \\
dicompyler \citep{Panchal2010} & 2009 & 2020-01 &  & 2 & 15941 & X & X &  &  \\
Fiji \citep{Schindelin2012} & 2011 & 2020-08 & X & 55 & 10833 & X & X & X &  \\
ImageJ \citep{Rueden2017} & 1997 & 2020-08 & X & 18 & 9681 & X & X & X &  \\
MRIcroGL \citep{Rorden2021} & 2015 & 2020-08 &  & 2 & 8493 & X & X & X &  \\
DicomBrowser \citep{Archie2012} & 2012 & 2020-08 &  & 3 & 5505 & X & X & X &  \\ \hline
\end{tabular}
\caption{\label{tab_final_list}Final software list}
\end{table}

Figure \ref{fig_language} shows the primary languages versus the number of
projects using them.  The primary language is the language used for the majority
of the project's code; in most cases projects also use other languages.  The
most popular language is C++, with almost 40\% of projects (11 of 29).  The two
least popular choices are Pascal and  Matlab, with around 3\% of projects each
(1 of 29).

\begin{figure}[ht]
\centering
\includegraphics{figures/PrimaryLanguages.pdf}
\centering
\caption{\label{fig_language}Primary languages versus number of projects}
\end{figure}

@@

Figure \ref{fg_installability_scores} lists the installability scores.  We found
installation instructions for 16 projects. Among the ones without instructions,
\textit{BioImage Suite Web} and \textit{Slice:Drop} do not need installation,
since they are web applications. Installing 10 of the projects required extra
dependencies. Five of them are the web applications (as shown in
Table~\ref{tab_final_list}) and depended on a browser; \textit{dwv},
\textit{OHIF Viewer}, and \textit{GATE} needed extra dependencies to build;
\textit{ImageJ} and	\textit{Fiji} needed an unzip tool; \textit{MatrixUser} was
based on Matlab; \textit{DICOM Viewer} needed to work on a Nextcloud platform.

@@

\textit{3D Slicer} has the highest score because it had easy to follow
installation instructions, and the installation processes were automated, fast,
and frustration-free, with all dependencies automatically added. There were also
no errors during the installation and uninstallation steps. Many other software
packages also had installation instructions and automated installers, and we had
no trouble installing them, such as \textit{INVESALIUS 3}, \textit{Gwyddion},
\textit{XMedCon}, and \textit{MicroView}. We gave them various scores based on
the understandability of the instructions, installation steps, and user
experience. Since \textit{BioImage Suite Web} and \textit{Slice:Drop} needed no
installation, we gave them higher scores. \textit{BioImage Suite Web} also
provided an option to download cache for offline usage, which was easy
to apply.
